### 2010-10-16 papers/BMD2010/PhysicalParametersPaper/paper.tex
found an error in the tor and com assigments to the wheel torsional inertias
leaving the library
rough draft ready

\begin{figure}[htbp]

@@

		\includegraphics[width=3in]{../../../images/YellowFrameTorsionalThird.jpg}
	\caption{The digital level was mounted to a straight edge that was aligned

@@

    decrease the oscillation.}
	\label{fig:level}
\begin{figure}[htbp]
        \includegraphics[width=3in]{../../../images/pendDist.jpg}
    \caption{Measuring the distance from the pendulum axis to the rear wheel
    axle using level ruler.}
    \label{fig:PendDist}

@@

perturb at small angles. The rate of oscillation was measured was measured

@@

Matlab. The measurement duration were either 15 or 30 secs and each moment of

@@

The physical phenomena that was observed for data sets such as these were that

@@

motion. Roland and Massing~\cite{Roland1971} noted that they had this same problem and solved it
by hold the torsional rod in a fixed bearing.

The period for a damped oscillation is
\begin{equation}
    T = \frac{2\pi}{\sqrt{1-\zeta^2}\omega_n}
    \label{eqn:periodDamped}
\end{equation}
and for light damping
\begin{equation}
    T \approx \frac{2\pi}{\omega_n}
    \label{eqn:period}
\end{equation}
The uncertainty in the period, $T$, can be determined from the fit. Firstly,
the variance of the fit is
\begin{equation}
    \sigma_y^2 =
    \frac{1}{N-5}\sum_{i=1}^N(y_{mi}-\bar{y}_m)^2-(y_{pi}-\bar{y}_m)^2
    \label{eqn:fitVariance}
\end{equation}
The covariance matrix of the fit function can be formed
\begin{equation}
    \mathbf{U} = \sigma_y^2\mathbf{H}^{-1}
    \label{eqn:covariance}
\end{equation}
where $\mathbf{H}$ is the Hessian~\cite{Hubbard1989b}. $\mathbf{U}$ is a $5\times5$ matrix with the
variances of each of the five fit parameters along the diagonal. This assumes
that the parameters are uncorrelated [note for Jason: you may have to take into
account correlation for this..maybe checkout the R function for nls]. The
variance of $T$ can be computed using the variance of $\zeta$ and $\omega$. It
is important to note that the uncertainties in the period are very low
($<10e-4$), even for the fits with low $r^2$ values due to the beating issue.
\subsection{Torsional Pendulum}
A torsional pendulum was used to measure all of moments of inertia about axes
in the lateral symmetric plane of each of the wheels, fork and frame. The
pendulum is made up of a rigid mount, an upper clamp, a torsion rod, and
various lower clamps.
\begin{figure}[htbp]
    \begin{center}
        \includegraphics[width=3in]{../../../images/fixture.jpg}
    \end{center}
    \caption{The rigid pendulumfixture mounted to a concrete column.}
    \label{fig:fixture}
\end{figure}
A 5mm diameter, 1m long mild steel rod was used as the torsion
spring. A lightweight, low relative moment of inertia clamp was constructed
that could clamp the rim and the tire. The wheel was hung freely such that the
center of mass aligned with the torsional pendulum axis and then it was
secured. The wheel was then perturbed and it oscillated about the pendulum axis.
The rate gyro was mounted on the clamp in line with the pendulum axis. The
radial moment of inertia can can calculated as such:
\begin{figure}[tbp]
    \begin{center}
        \includegraphics[width=4in]{../../../images/rod.jpg}
    \end{center}
    \caption{The steel calibration rod.}
    \label{fig:rod}
\end{figure}
The torsional pendulum was calibrated using a known moment of inertia.
A torsional pendulum almost identical to the one used in
\cite{Kooijman2006} was used to measure the averaged period $\overline{T}_i$ of
oscillation of the rear frame at three different
orientation angles $\beta_i$, where $i=1$, $2$, $3$, as shown in
Fig.~\ref{fig:triangle}. The parts were perturbed lightly, less than 1 degree,
and allowed to oscillate about the pendulum axis through at least ten periods.
The time of oscillation was recorded via a stopwatch ($\pm1$~s). This was done
three times for each frame and the recorded times were averaged. The
coefficient of elasticity $k$ for the torsional pendulum had previously been
measured in \cite{Kooijman2006} and found to be $k=5.01\pm0.01$
$\frac{\textrm{Nm}}{\textrm{rad}}$.

\subsection{Wheels}
The wheels are assumed to be symmetric laterally and about any radial axis. Thus only two
moment's of inertia are required. The moment of inertia about the axle was
measured by hanging the wheel as a compound pendulum. The wheel was hung on a
horizontal rod and perturbed to oscillate about the axis of the rod. This rate
gyro was attached to the spokes near the hub and orientated mostly along the
axle axis. The wheels tended to precess at the contact point about the vertical
axis. This adds a very low frequency component of rate along the vertical axis,
but should not affect the compound pendulum axis. A better fixture for the
wheel could prevent the precession. The pendulum arm length is the distance
from the rod/rim contact point to the mass center of the wheel. The inner
diameter of the rim was measured and divided by two to get $l$. The moment of
inertia about the axle is calculated by:
\begin{equation}
    I_{\mathrm{R}yy} = \left(\frac{\bar{T}}{2\pi}\right)^2m_\mathrm{R}gl_\mathrm{R} -
    m_\mathrm{R}l^2
\end{equation}
\begin{figure}[tbp]
    \centering
    \subfloat[]{
        \label{fig:FwheelTor}
        \includegraphics[width=2.75in]{../../../images/CrescendoFwheelTorsionalFirst.jpg}
    }
    \quad
    \subfloat[]{
        \label{fig:wheelIyy}
        \includegraphics[width=2.75in]{../../../images/wheelIyy.jpg}
    }
    \caption{\subref{fig:FwheelTor} The front wheel of the Crescendo hung as a
    torsional pendulum. \subref{fig:wheelIyy} A wheel hung as a compound pendulum.}
    \label{fig:wheelPend}
\end{figure}
The radial moment of inertia was measured by hanging the wheel as a torsional
pendulum. A 5mm diameter, 1m long mild steel rod was used as the torsion
spring. A lightweight, low relative moment of inertia clamp was constructed
that could clamp the rim and the tire. The wheel was hung freely such that the
center of mass aligned with the torsional pendulum axis and then it was
secured. The wheel was then perturbed and it oscillated about the pendulum axis.
The rate gyro was mounted on the clamp in line with the pendulum axis. The
radial moment of inertia can can calculated as such:
\begin{equation}
    I_{xx} = \frac{k\bar{T}^2}{4\pi^2}
\end{equation}
Finding the inertia tensors of the wheels is less complex because the wheels
are symmetric about three orthogonal planes so there are no products of
inertia. The $I_{xx}=I_{zz}$ moments of inertia were calculated by measuring
the averaged period of oscillation about an axis in the $XZ$-plane using the
torsional pendulum setup and Eq.~\ref{eq:torPend}. The $I_{yy}$ moment of
inertia was calculated with a compound pendulum as described in
\cite{Kooijman2006} and shown in Fig. using
\begin{equation}
	I_{yy}=\left(\frac{\overline{T}}{2\pi}\right)^2mgl-ml^2
\label{eq:comPend}
\end{equation}
where $l=0.303\pm0.002$ m is the pendulum length, $m$ is the mass of the wheel,
$\overline{T}$ is the averaged period and $g$ is the local acceleration due to
gravity. Table  gives the calculated values.
\subsection{Frame}
Three measurements were made to estimate the globally referenced moments and
products of inertia ($I_{xx}$, $I_{xz}$ and $I_{zz}$) of the rear frame. The
frame was typically hung from the three main tubes: seat tube, down tube
and top tube. The rear fender prevented easy connection to the seat tube on
some of the bikes and the clamp was attached to the fender. The fender was
generally less rigid than the frame tube. For best accuracy with only three orientation angles, the frame
should be hung at three angles that are $120^\circ$ apart. The three tubes on
the frame generally provide that the orientation angles were spread evenly at
about $120^\circ$. Taking data at more orientation angles could improve the accuracy further and
is gernally possible with standard diamond frame bicylces.

Three moments of inertia $J_{i}$ about the pendulum axes were calculated with
\begin{equation}
	J_i=\frac{k\overline{T}_i^2}{4\pi^2}
\label{eq:torPend}
\end{equation}

@@

	\mathbf{J}_i=\mathbf{R}_i^T\mathbf{IR}_i

@@

$\mathbf{R}$ is the rotation matrix relating the two frames. The global inertia

@@

		I_{xx}  & -I_{xz}\\
		-I_{xz} & I_{zz}

@@

the same as the $y$ axis of the benchmark reference frame. The simple rotation matrix about the $Y$-axis

@@

		c_{\beta i} & s_{\beta i}\\
		-s_{\beta i} & c_{\beta i}

@@

The first entry of $\mathbf{J}_i$ in Eq.~\ref{eq:rotIn} is the moment of inertia about the pendulum axis and is written explicitly as 

@@

	J_{i}=c^{2}_{\beta i}I_{xx}+2s_{\beta i}c_{\beta i}I_{xz}+s^{2}_{\beta i}I_{zz}\textrm{.}

@@

@@

@@

    \includegraphics[width=4in]{../../../plots/Bike/eig_plot.pdf}

@@

        \includegraphics[width=4in]{../../../plots/Bike/CrescendoRootLoci.pdf}

@@

@@

    \includegraphics[width=4in]{../../../plots/BikeRider/eig_plot.pdf}

@@

torque).

@@

    \includegraphics[width=4in]{../../../plots/Bike/Bode/Tdel2phi.pdf}

@@

    \includegraphics[width=4in]{../../../plots/BikeRider/Bode/Tdel2phi.pdf}

@@

\bibliographystyle{acm}
\bibliography{bicycle}
\appendix
\section{RAW DATA}
%\begin{landscape}
%\subfile{../../../tables/Bike/Parameters/MasterParTable.tex}
%\end{landscape}
\section{PARAMETERS}
\subsection{BICYCLE}
\begin{landscape}
\subfile{../../../tables/Bike/Parameters/MasterParTable.tex}
\end{landscape}
\subsection{BICYCLE WITH RIDER}
%\begin{landscape}
%\subfile{../../../tables/Bike/Parameters/MasterParTable.tex}
%\end{landscape}
\section{CANONICAL MATRICES}
\subsection{BICYCLE}
\begin{landscape}
\subfile{../../../tables/Bike/Canonical/MasterCanTable.tex}
\end{landscape}
\subsection{BICYCLE WITH RIDER}
%\begin{landscape}
%\subfile{../../../tables/Bike/Parameters/MasterParTable.tex}
%\end{landscape}
\end{document}

>>>>

@@

    \subfloat[]{\label{fig:level}\includegraphics[width=2.75in]{../../../images/YellowFrameTorsionalThird.jpg}}
    \quad
    \subfloat[]{\label{fig:PendDist}\includegraphics[width=2.75in]{../../../images/pendDist.jpg}}
    \caption{\subref{fig:level} The digital level was mounted to a straight edge that was aligned

@@

    decrease the oscillation. \subref{fig:PendDist} Measuring the distance from
    the pendulum axis to the rear wheel axle using level ruler.}

@@

perturb at small angles. The rate of oscillation was measured

@@

\textsc{Matlab}. The measurement duration were either 15 or 30 secs and each moment of

@@

The physical phenomena observed for corresponding to data sets such as these were that

@@

motion.

The period for a damped oscillation is
\begin{equation}
    T = \frac{2\pi}{\sqrt{1-\zeta^2}\omega_n}
    \label{eqn:periodDamped}
\end{equation}
and for light damping
\begin{equation}
    T \approx \frac{2\pi}{\omega_n}
    \label{eqn:period}
\end{equation}
The uncertainty in the period, $T$, can be determined from the fit. Firstly,
the variance of the fit is
\begin{equation}
    \sigma_y^2 =
    \frac{1}{N-5}\sum_{i=1}^N(y_{mi}-\bar{y}_m)^2-(y_{pi}-\bar{y}_m)^2
    \label{eqn:fitVariance}
\end{equation}
The covariance matrix of the fit function can be formed
\begin{equation}
    \mathbf{U} = \sigma_y^2\mathbf{H}^{-1}
    \label{eqn:covariance}
\end{equation}
where $\mathbf{H}$ is the Hessian~\cite{Hubbard1989b}. $\mathbf{U}$ is a $5\times5$ matrix with the
variances of each of the five fit parameters along the diagonal.
The variance of $T$ can be computed using the variance of $\zeta$ and $\omega$. It
is important to note that the uncertainties in the period are very low
($<10e-4$), even for the fits with low $r^2$ values.
\subsection{TORSIONAL PENDULUM}
A torsional pendulum was used to measure all of moments of inertia about axes
in the lateral symmetric plane of each of the wheels, fork and frame. The
pendulum is made up of a rigid mount, an upper clamp, a torsion rod, and
various lower clamps.
\begin{figure}[tbp]
    \begin{center}
        \includegraphics[width=2in]{../../../images/fixture.jpg}
    \end{center}
    \caption{The rigid pendulum fixture mounted to a concrete column.}
    \label{fig:fixture}
\end{figure}
A 5 mm diameter, 1 m long mild steel rod was used as the torsion
spring. A lightweight, low relative moment of inertia clamp was constructed
that could clamp the rim and the tire. The moments of inertia of the clamps
were neglected. The wheel was hung freely such that the
center of mass aligned with the torsional pendulum axis and then it was
secured. The wheel was then perturbed and it oscillated about the pendulum
axis. The rate gyro was mounted on the clamp orientated along the pendulum
axis.

The torsional pendulum was calibrated using a known moment of
inertia~\ref{fig:rod}. A torsional pendulum almost identical to the one used in
\cite{Kooijman2006} was used to measure the averaged period $\overline{T}_i$ of
oscillation of the rear frame at three different
orientation angles $\beta_i$, where $i=1$, $2$, $3$, as shown in
Fig.~\ref{fig:triangle}. The parts were perturbed lightly, less than 1 degree,
and allowed to oscillate about the pendulum axis through at least ten periods.
This was done at least three times for each frame and the recorded periods were
averaged.
        \includegraphics[width=4in]{../../../images/rod.jpg}
    \caption{The steel calibration rod. The moment of inertia of the rod,
    $I=\frac{m}{12}(3r^2+l^2)$, can be used to estimate the stiffness of the
    pendulum, $k=\frac{4I\pi^2}{\overline{T}^2}$, with $k=5.62\pm0.02$ $\frac{\textrm{Nm}}{\textrm{rad}}$}
    \label{fig:rod}
Finding the full inertia tensors of the wheels is less complex because the wheels
are assumed symmetric about three orthogonal planes so there are no products of
inertia. The $I_{xx}=I_{zz}$ moments of inertia were calculated by measuring
the averaged period of oscillation about an axis in the $XZ$-plane using the
torsional pendulum setup and Eq.~\ref{eq:torPend}.The wheels are assumed to be symmetric laterally and about any radial axis. Thus only two
moment's of inertia are required for the benchmark parameters set. The moment
of inertia about the axle was measured by hanging the wheel as a compound
pendulum, Fig.~\ref{fig:wheelIyy}. The wheel was hung on a
horizontal rod and perturbed to oscillate about the axis of the rod. This rate
gyro was attached to the spokes near the hub and orientated mostly along the
axle axis. The wheels tended to precess at the contact point about the vertical
axis whixh added a very low frequency component of rate along the vertical
radial axis, but should not affect the period estimation about the compound
pendulum axis. A fixture that prevented the precession would be preferable for
future measurements. The pendulum arm length is the distance
from the rod/rim contact point to the mass center of the wheel. The inner
diameter of the rim was measured and divided by two to get $l_\mathrm{F,R}$. The moment of
inertia about the axle is calculated by:
    I_{\mathrm{R}yy} = \left(\frac{\bar{T}}{2\pi}\right)^2m_\mathrm{R}gl_\mathrm{R} -
    m_\mathrm{R}l^2
    \label{eq:CompoundInertia}
    need the sigma for IRyy
        \label{fig:FwheelTor}
        \includegraphics[width=2.75in]{../../../images/CrescendoFwheelTorsionalFirst.jpg}
    \quad
        \label{fig:wheelIyy}
        \includegraphics[width=2.75in]{../../../images/wheelIyy.jpg}
    \caption{\subref{fig:FwheelTor} The front wheel of the Crescendo hung as a
    torsional pendulum. \subref{fig:wheelIyy} A wheel hung as a compound pendulum.}
    \label{fig:wheelPend}
The radial moment of inertia was measured by hanging the wheel as a torsional
pendulum, Fig.~\ref{fig:FwheelTor}. The wheel was hung freely such that the center of mass aligned with
the torsional pendulum axis and then it was secured in position. The wheel was then
perturbed and it oscillated about the pendulum axis. The radial moment of inertia can can calculated as such:
    I_{xx} = \frac{k\bar{T}^2}{4\pi^2}
    need the sigma for Ixx
\subsection{FRAME}
Three measurements were made to estimate the globally referenced moments and
products of inertia ($I_{xx}$, $I_{xz}$ and $I_{zz}$) of the rear frame. The
frame was typically hung from the three main tubes: seat tube, down tube
and top tube, Fig.~\ref{fig:level}. The rear fender prevented easy connection to the seat tube on
some of the bikes and the clamp was attached to the fender. The fender was
generally less rigid than the frame tube. For best accuracy with only three
orientation angles, the frame should be hung at three angles that are
$120^\circ$ apart. The three tubes on the frame generally provide that the
orientation angles were spread evenly at about $120^\circ$. Furthermore, taking
data at more orientation angles could improve the accuracy further and
is generally possible with standard diamond frame bicylces.

Three moments of inertia $J_{i}$ about the pendulum axes were calculated using
Eq.~\ref{eq:torPend}.
	J_i=\frac{k\overline{T}_i^2}{4\pi^2}
\label{eq:torPend}
    need sigma

@@

	\mathbf{J}_i=\mathbf{R}_i\mathbf{IR}_i^T

@@

$\mathbf{R}$ is the rotation matrix relating the two frames,
Fig.~\ref{fig:angles}. The global inertia

@@

		I_{xx}  & I_{xz}\\
		I_{xz} & I_{zz}

@@

the same as the $y$ axis of the benchmark reference frame. The simple rotation
matrix about the $Y$-axis

@@

		c_{\beta i} & -s_{\beta i}\\
		s_{\beta i} & c_{\beta i}

@@

The first entry of $\mathbf{J}_i$ in Eq.~\ref{eq:rotIn} is the moment of
inertia about the pendulum axis and is written explicitly as

@@

	J_{i}=c^{2}_{\beta i}I_{xx}-2s_{\beta i}c_{\beta i}I_{xz}+s^{2}_{\beta i}I_{zz}\textrm{.}

@@

The inertia of the frame about an axis normal to the plane of symmetry was
estimated by hanging the frame as a compound pendulum at the wheel axis,
Fig.~\ref{fig:frameCompound}. Equation~\ref{eq:CompoundInertia} is used but
with the mass of the frame and the frame pendulum length.
    l_B=\sqrt{x_B^2+(z_B+r_R)^2}
    \label{eq:FramePendLength}

@@

\subsection{FORK AND HANDLEBAR}
The inertia of the fork and handlebar is calculated in the same way as the frame. The fork is
hung as both a torsional pendulum, Fig.~\ref{fig:StratosFork}, and as a compound pendulum,
Fig.~\ref{fig:forkCompound}. The fork provides fewer mounting options to
obtain at least three equally spaced orientation angles, especially if there is
no fender. The torsional calculations follow equations~\ref{eq:torPend} through
\ref{eq:inRel} and the compound is calculated with
equation~\ref{eq:CompoundInertia}. The fork pendulum length is calculated using
    l_H=\sqrt{(x_H-w)^2+(z_B+rF)^2}

@@

    \includegraphics[width=6in]{../../../plots/Bike/eig_plot.pdf}

@@

        \includegraphics[width=6in]{../../../plots/Bike/CrescendoRootLoci.pdf}

@@

    \caption{The eigenvector components for the mode with the second complex
    pair at low speeds.}

@@

    \includegraphics[width=6in]{../../../plots/BikeRider/eig_plot.pdf}

@@

torque). The difference in the frequency response for the bicycle with the
rigid rider shows less variation among the bicycles,
Fig.~\ref{fig:bikeRiderBode}, as the rider's mass and
inertia play a bigger roll.

@@

    \includegraphics[width=6in]{../../../plots/Bike/Bode/Tdel2phi.pdf}

@@

    \includegraphics[width=6in]{../../../plots/BikeRider/Bode/Tdel2phi.pdf}

@@

\section{CONCLUSION}
We present a detailed method of accurately estimating the physical parameters
of a bicycle need for the benchmarked Whipple bicycle
model~\cite{Meijaard2007}. Then we measure eight different bicycles providing
both the raw data, parameter sets and linear model coefficient matrices with
the uncertainties for each value. Then we present a brief comparison of the
eight bicycles using eigenanalysis and Bode analysis.

\bibliographystyle{acm}
\bibliography{bicycle}
\appendix
%\section{RAW DATA}
%\begin{landscape}
%\subfile{../../../tables/Bike/Parameters/MasterParTable.tex}
%\end{landscape}
%\section{PARAMETERS}
%\subsection{BICYCLE}
\begin{landscape}
\subfile{../../../tables/Bike/Parameters/MasterParTable.tex}
%\end{landscape}
%\subsection{BICYCLE WITH RIDER}
%\begin{landscape}
\subfile{../../../tables/BikeRider/Parameters/MasterParTable.tex}
%\end{landscape}
%\section{CANONICAL MATRICES}
%\subsection{BICYCLE}
%\begin{landscape}
\subfile{../../../tables/Bike/Canonical/MasterCanTable.tex}
%\end{landscape}
%\subsection{BICYCLE WITH RIDER}
%\begin{landscape}
\subfile{../../../tables/BikeRider/Canonical/MasterCanTable.tex}
\end{landscape}
\end{document}
